Biodiversity temporal trends are reshaping food web structure and redundancy in riverine ecosystems

### 2025-02-07 paper/manuscript.Rmd
Improve hierarchy and clarity of the objectives and findings:
- Improve hierarchy
- Mirror discussion and intro
- Mv details of spatial analysis to appendix

---
title: "Biodiversity changes are linked to changes in food-web structure and redundancy in riverine ecosystems"
subtitle: "Alain Danet (a,b), Loubna El Madouri (a,c), Andrew P. Beckerman (b),
Maud Mouchet (a), Colin Fontaine (a) and Elisa Thébault (d)"
date: "Research report"
output:
  officedown::rdocx_document:
    number_sections: false
    base_format: "bookdown::word_document2"
always_allow_html: true
number_sections: false
bibliography: references.bib
link-citations: true
#csl: ecology-letters.csl
toc: false
linestretch: 2.0
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{hyperref}
  - \hypersetup{backref=true}
  - \usepackage{setspace}\doublespacing
csl: pnas.csl
---

@@

(a) Centre d’Ecologie et des Sciences de la Conservation, UMR 7204 MNHN7
CNRS-Sorbonne Université, Muséum national d’Histoire naturelle de Paris, 43 rue
8 Buffon, 75005 Paris, France

@@

# Abstract (152 / 250 words max)

Changes in biodiversity can fundamentally alter ecological communities, yet how
these changes affect food web structure remains poorly understood. Here, we
analyze temporal trends in over 400 riverine fish communities across France
spanning 1995-2018, examining relationships between species richness, community
biomass, and three key metrics of food web structure: connectance, weighted
average trophic level, and trophic pathway redundancy. Using structural equation
modeling, theoretical simulations, and robustness analyses, we demonstrate that declining
community biomass leads to loss of top trophic levels and decreased food web
connectance, while species richness negatively correlates with connectance but
positively affects trophic redundancy. Notably, both biomass and species
richness positively influence the proportion of redundant trophic links,
suggesting that biodiversity loss reduces community robustness to future
perturbations. Our findings, consistent across temporal, spatial, and
theoretical approaches, reveal how biodiversity change restructures food webs
and highlights the utility of food web metrics as indicators of ecosystem
function and fragility.

# Significance (124 / 120 words max)

Understanding how biodiversity loss affects ecological communities is crucial
for predicting and mitigating human impacts on ecosystems. This study
provides comprehensive evidence linking biodiversity changes to food web
structure using an extensive dataset of riverine fish communities. We
demonstrate that changes in community biomass and species richness
systematically alter food web architecture, affecting trophic levels,
connectance, and the redundancy of feeding relationships. Our findings unite
multiple ecological theories, showing how energetic constraints and
biodiversity-ecosystem functioning mechanisms jointly shape food web assembly
and disassembly. Importantly, we show that declining biodiversity reduces the
redundancy of trophic pathways, potentially making communities more vulnerable
to future perturbations. These results suggest that food web structure metrics
could serve as valuable indicators for monitoring ecosystem health and
evaluating conservation strategies.

@@

Ecological communities are experiencing massive changes due to human pressures,
which are expected to have striking consequences on ecosystem functioning and
fragility. It is increasingly recognized that such changes across space are
associated with spatial variations in human pressures [@mcgill_fifteen_2015;
@blowes_geography_2019], such as freshwater
and terrestrial biodiversity declines linked to land degradation
[@newbold_global_2015; @cardinale_is_2018; @danet_past_2024], and declining
biomass of marine fishes linked to increased fishing pressures on predatory
fishes [@christensen_century_2014]. These documented changes in diversity and
biomass result in altered community structure and composition
[@dornelas_assemblage_2014; @hillebrand_biodiversity_2018; @danet_past_2024].

The structure of ecological networks, which describe the number, distribution
and strength of interactions among species, is important because it, along with
species richness per se, underpins species coexistence, ecosystem functioning
and the stability of ecological dynamics and ecosystem function
[@chesson_interaction_2008; @thebault_food-web_2003; @oconnor_general_2017;
@danet_species_2021; @eschenbrenner_diversity_2023; @danet_response_2024].
Theory and empirical data have shown that species richness and food-web
structure impact on coexistence [@chesson_interaction_2008], ecosystem
functioning [@thebault_food-web_2003; @wang_biodiversity_2018], stability
[@danet_species_2021; @danet_response_2024; @eschenbrenner_diversity_2023], and
ecosystem robustness to perturbations [@dunne_network_2002;
@allesina_functional_2009; @bodini_using_2009; @allesina_functional_2009].
Keystone predator species regulating dominant competitors can limit competitive
exclusion [@heithaus_predicting_2008; @johnson_rarity_2007;
@estes_trophic_2011], and thereby maintain species richness. Food-webs with
higher trophic levels are linked to elevated  ecosystem productivity
[@thebault_food-web_2003; @wang_biodiversity_2018] and community temporal
stability [@danet_species_2021; @eschenbrenner_diversity_2023;
@danet_response_2024]. Finally, the number and distribution of trophic links in
food-webs determine the robustness of communities to perturbations
[@dunne_network_2002; @allesina_functional_2009; @bodini_using_2009;
@allesina_functional_2009]. A better knowledge of the consequences of
biodiversity change on the structure of ecological networks is thus necessary to
quantify and mitigate anthropogenic driven changes to biodiversity.

There are three key components to the mapping between biodiversity changes and
food-web structure. First, the energetic paradigm underpinning the trophic
theory of biogeography [@lindeman_trophic-dynamic_1942; @gravel_trophic_2011]
posits that the presence of top consumers requires resources in enough numbers
at lower trophic levels derived from bottom up productivity. From this, it can
be further hypothesized that the presence of generalist top consumers will be
strongly conditioned by the amount of community biomass.

Second, connectance (the proportion of realized trophic links), is commonly
reported to decrease as species richness increases [@oconnor_general_2017;
@dunne_network_2006], because consumers are limited in the number of resources
they can consume. Connectance has also been linked to the robustness to species
extinctions [@dunne_network_2002; @allesina_who_2004; @bodini_using_2009]. More
in detail, it is the prevalence of redundant (versus functional) links in food
webs [@allesina_who_2004; @bodini_using_2009; @allesina_functional_2009], i.e.
the redundancy of trophic pathways, that predicts ecosystem robustness as they
can be safely removed without triggering secondary extinctions. We thus predict
that higher species richness covaries with higher redundancy of trophic pathways
in food-webs [@allesina_functional_2009]. Together, these relationships mean
that biodiversity changes are likely to be associated with changes in trophic
links distribution and then possibly to changes in robustness to perturbations,
such as increases in species richness should increase the number of redundant
trophic pathways to top predators and then increase community robustness.

Finally, biodiversity and ecosystem functioning theory suggest specifically that
community biomass should be positively related to species richness
[@loreau_biodiversity_2001]. This arises specifically because of niche
complementarity where species exploit different resources. In the particular
case of food webs, theoretical studies show that the presence of higher trophic
level species decreases niche overlap among producers which then increase
productivity and community biomass [@schneider_animal_2016;
@wang_biodiversity_2018; @wang_intraguild_2019; @nie_will_2023].

Here, we report on an analysis of riverine fish community times series to
document the ecological consequences of biodiversity changes on food web
structure. We specifically aim to reveal the relationship between community
biomass and species richness and three key metrics of food web structure:
connectance, weighted average trophic level, and the proportion of redundant
trophic links (hereafter “redundancy”). To do this, we implemented a sequence of
analyses that allows us to enumerate these relationships as covariation among
temporal trends. First, we estimated `r nb_op` food web structure for `r nb_st`
fish communities with 10+ years of community dynamics spanning 1995-2018 over
France (Fig. \@ref(fig:matmet)a-c, Table S1).  Second, using these food webs, we
estimated the temporal trends in each of species richness, biomass, connectance,
trophic structure and redundancy. Having estimated each temporal trend, we next
constructed an SEM to make specific inference about the relative effects of
community biomass and species richness  temporal trends on the temporal trends
three metrics of food web structure, and the effect of species richness on
community biomass (Fig. \@ref(fig:matmet)d). We also implemented a robustness analysis on the
meta-food web (i.e. the food web grouping all trophic species found in the
dataset and their inferred interactions) to identify species extinction
scenarios that might explain the variation in food web structure that we see in
the timeseries. Finally, we performed three further analysis to evaluate the
robustness and the generality of our results. We replicated our temporal
analysis with a space-for-time approach (Fig. \@ref(fig:matmet)c & e) using median metrics in
each site instead of the temporal trends; we created a simulation of complex
communities using a bioenergetic food web model to assess the generality of our
results in complex communities that allow relaxation of some of the constraints
defined by our fish biodiversity data; and we performed a three-scenario,
topological secondary extinction analysis on the full empirical metaweb to
evaluate the role of top predators on food-web structure.

```{r}
fig_mat_cap <- paste0("Using ", nb_st," sites of stream fishes in (a) French streams that
  have been monitored at least 10 years, (b) we reconstructed the food-web at each
  site and each sampling event by building a metaweb based on ontogenic food diet information for each
  species and predator-prey body size ratio. (c) We computed the temporal trends
  and the median of community biomass, species richness, connectance, average
  trophic level, and redundancy at each site. We then assessed the relationships
  among food web structure descriptors, (d) in time and (e) across
  space. We additionally performed extinction simulations from the metaweb and
  insilico experiments using a bioenergetic food web model to gain more insights
  and assess the generality of our results (see Methods).")
```

@@

Our food web reconstruction delivered `r nb_op` estimates of species richness,
biomass, connectance, weighted average trophic level, and the proportion of
redundant trophic links (hereafter "redundancy", see Appendix B) from our 403
fish communities with 10+ years of population dynamics spanning 1995-2018 over
France (Fig. 1a-c). We implemented Bayesian hierarchical models to jointly
estimate the temporal trends in these metrics. We found a great heterogeneity in
the temporal trends at the site level (community biomass: from `r grbmn`% to
`r grbmx`% per decade, species richness: `r grrhn` to +`r grrhx`%, average
trophic level: `r grtln` to +`r grtlx`, connectance: `r grctn` to +`r grctx`,
redundancy: `r grln` to +`r grlx`, Fig. S1) which was effectively captured by
the hierarchical model (with conditional R2 between `r r2cond$log_bm_std` for
community biomass to `r r2cond$log_rich_std` for species richness, Table S2).

```{r}
sem_cap <- "(a) Structural equation models linking the temporal trends of the
food-web structure. The path widths are proportional to the standardized slope
coefficients. The standardized slope coefficients and their associated 95%
  confidence intervals obtained by bootstrapping are displayed along the path.
R2 displays the marginal R-squared of the linear models. (b to h) Direct
bivariate relationships derived from the structural equation model. Points
display the temporal trends at each site estimated with a hierarchical Bayesian
model, lines display the linear relationship estimated with the SEM. The
distribution of the site temporal trends are displayed in Fig. S1. All the
effects are reported in Table S3. Average trophic level is weighted by the
biomass of the trophic species. Redundancy is the proportion of redundant
links."
```

@@

## Association between biodiversity and food-web temporal trends

These temporal trend estimations at the site level were the input to our SEM,
allowing us to evaluate the direct and indirect effect of species richness and
community biomass changes on food web structure changes and between species
richness and community biomass. We detected a strong positive association
between community biomass and in species richness ($r_\delta =$ `r rdtps$bm$r`,
Fig. \@ref(fig:sem) a and b, $r_\delta$: standardized slope coefficient, [95%
CI], Table S3). This indicates that communities with decreasing species richness
over time experienced a concomitant decrease in community biomass and
vice-versa.

Our SEM also predicts a positive effect of community biomass on connectance,
average trophic level, and redundancy (resp. $r_\delta =$ `r rdtps$ct$bm`,
`r rdtps$tlvl$bm`, `r rdtps$red$bm`, Fig. \@ref(fig:sem)b, c, d). Thus, when communities
experienced a decrease in biomass, they also experimented a concomitant decrease
in connectance, average trophic level and redundancy, and vice versa. We further
found that the temporal trends in community biomass were positively associated
with both the number and the proportion of piscivorous species (Fig. S2a),
suggesting that the presence of high trophic level trophic species has a
disproportionate role community biomass.

The effects of species richness were more complex, because species richness in
this model has indirect effects on food-web structure through community biomass.
Species richness had a direct negative effect on connectance ($r_\delta =$
`r rdtps$ct$r`, Fig. \@ref(fig:sem)f), a positive indirect effect through
community biomass ($r_\delta =$ `r ridtps$ct$r`), so that species richness had a
negative effect in total on connectance ($r_\delta =$ `r rttps$ct$r`, Fig.
\@ref(fig:std)a). In contrast, species richness had no direct effects on in
average trophic level ($r_\delta =$ `r rdtps$tlvl$r`, Fig. \@ref(fig:sem)g), a
positive indirect effect through community biomass ($r_\delta =$
`r ridtps$tlvl$r`), which adds up to a positive total effect of species richness
temporal trends on average trophic level ($r_\delta =$ `r rttps$tlvl$r`, Fig.
\@ref(fig:std)a). Species richness temporal trends had direct and indirect
positive effect on the temporal trends in redundancy (resp. $r_\delta =$
`r rdtps$red$r`, `r ridtps$red$r`), which added up to a strong positive effect
($r_\delta =$ `r rttps$red$r`).

Furthermore, we found that temporal trends of the proportion of piscivorous
species was negatively associated with the temporal trends of species richness
(Fig. S2), meaning that communities that increase in richness over time have an
increase in the proportion of non-piscivorous species over time and vice versa.
It means food-webs that have experienced declines in species richness over the
last decades have decreased in average trophic level and redundancy but increase
in connectance, and conversely, those that experienced increases in species
richness have increased in average trophic level, redundancy, but decreased in
connectance.

## Theoretical simulations and space-for-time analysis (mostly) agree with the temporal analysis

We further assessed the consistency of the reported associations among temporal
trends by implementing two additional analyses. We first tested if the
associations among the temporal trends within sites was holding across sites,
i.e. using spatial variations in food-web structure rather than temporal changes
over time (space-for-time hypothesis). Second, we performed theoretical
simulations with a bioenergetic food-web model to cross-reference our site and
species specific empirical data on riverine fish with a more generalized
representation of food web dynamics known to represent accurate relationships
among our focal variables [@brose_allometric_2006; @eschenbrenner_diversity_2023;
@danet_response_2024]. With those additional empirical and theoretical data, we
assessed additional SEM models with the same structure as in Section 4.2.

As in the temporal analysis, species richness had a positive effect on community
biomass both in empirical data and in theoretical simulations (resp. $r_\delta
=$ `r rdsp$bm$r` and `r rdm$bm$r`, Fig. \@ref(fig:std)a and S3). In both
empirical data and theoretical simulations, community biomass had a positive
effect on average trophic level (resp. $r_\delta =$ `r rdsp$tlvl$bm` and
`r rdm$tlvl$bm`), and redundancy (resp. $r_\delta =$ `r rdsp$red$bm` and
`r rdm$red$bm`), as in temporal analysis. We found weak evidence of a positive
effect of spatial variation in community biomass on connectance in empirical
data ($r_\delta =$ `r rdsp$ct$bm`) but strong evidence in theoretical
simulations ($r_\delta =$ `r rdm$ct$bm`, Fig. \@ref(fig:std)a).

The effects of spatial variations in species richness on food-web structure
mostly aligned with those from the temporal trend analysis, except for average
trophic level. We found a direct negative effect of species richness on
connectance both using data and model simulations (resp. $r_\delta =$
`r rdsp$ct$r` and `r rdm$ct$r`, Fig. \@ref(fig:std)a), on average trophic level
(resp. $r_\delta =$ `r rdsp$tlvl$r` and `r rdm$tlvl$r`), but a direct positive
effect on redundancy (resp. $r_\delta =$ `r rdsp$red$r` and `r rdm$red$r`).
Species richness had an indirect positive effect of species richness through
community biomass, which added up to strong total negative effects on
connectance in empirical data and simulations (resp. $r_\delta =$
`r rtsp$ct$r` and `r rtm$ct$r`) and positive on redundancy (resp. $r_\delta =$
`r rtsp$red$r` and `r rtm$red$r`). We found that
species richness had a total positive effect on average trophic level in
bioenergetic model, in agreement with temporal trends (resp. $r_\delta =$ `r rtm$tlvl$r`),
but a total negative effect in spatial variations (resp. $r_\delta =$ `r rtsp$tlvl$r`).

@@

We implemented three secondary extinction simulation scenarios on the empirical
metaweb defined as a food-web containing all the trophic species found across
all sites in the empirical data. This was to access a second line of inference
about whether top species were drivers of biomass and structure patterns (see
section 4.2). When species are removed in a sequential order beginning by the
highest trophic levels, connectance, average trophic level, and redundancy
decreased (purple lines, Fig. \@ref(fig:std)b). On the other hand, removing low trophic level
species first increases connectance and average trophic level, but do not affect
redundancy (yellow lines, Fig. \@ref(fig:std)b). In turn, random species extinction sequence
had little effect on food-web structure (black line, Fig. \@ref(fig:std)b). These results, in
tandem with our empirical analyses above, suggest that top consumers play a
disproportionate role in driving the changes in food web structure that we
report.

```{r}
std_cap <- paste0( "a, Direct and total effects of community biomass and species
  richness trends on food-web structure trends. Standardized coefficients (r)
  derived from the temporal, spatial, and bioenergetic model Structural Equation
  Models. Points and lines display the estimated effects and their confidence
  intervals at 95\\%. Values are displayed in Table S3, R-squared in Table S4, and
  the visualization of the bivariate relationships for the spatial and
  bioenergetic model SEMs are displayed in Figure S3. Average trophic level is
  weighted by the biomass of the trophic species. Redundancy is the proportion
  of redundant links. b, Effect of species extinctions on the structure of the
  empirical metaweb according to three scenarios. In the Decreasing and
  Increasing scenarios, species were ordered by their trophic level and removed
  from the metaweb starting respectively from the highest and lowest trophic
  level. Species were removed in a random sequence (50 sequences, grey dots) for
  the random scenario, the black line displaying the average predicted by a GAM
  linear model. For the extinction sequence, the average trophic level was not
  weighted by the biomass of species as this information is not available at the
  metaweb scale. Similar results were obtained with trophic species (i.e. node)
  extinction (Fig. S4).")
```

>>>>

---
title: "Biodiversity temporal trends are reshaping food web structure and redundancy in riverine ecosystems"
subtitle: "Alain Danet (a,b), Loubna El Madouri (a,c), Andrew P. Beckerman (b),
Maud Mouchet (a), Colin Fontaine (a) and Elisa Thébault (d)"
date: "Research report"
output:
  officedown::rdocx_document:
    number_sections: false
    base_format: "bookdown::word_document2"
always_allow_html: true
number_sections: false
bibliography: references.bib
link-citations: true
#csl: ecology-letters.csl
#https://pkgs.rstudio.com/rticles/articles/examples.html
toc: false
linestretch: 2.0
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{hyperref}
  - \hypersetup{backref=true}
  - \usepackage{setspace}\doublespacing
csl: pnas.csl
---

@@

(a) Centre d’Ecologie et des Sciences de la Conservation, UMR 7204
MNHN-CNRS-Sorbonne Université, Muséum national d’Histoire naturelle de Paris, 43
rue Buffon, 75005 Paris, France

@@

# Abstract (229 / 250 words max)

Biodiversity changes are expected to profoundly impact ecosystem functioning and
stability, yet empirical evidence remains limited. Bridging this gap requires
aligning biodiversity trends with food web theory and Biodiversity-Ecosystem
Functioning research, which provide strong predictions about the ecological
consequences of biodiversity loss. Most insights rely on spatial comparisons of
ecological communities, assuming they reflect temporal trends—a space-for-time
substitution. Here, we analyze biodiversity and food web time series from over
400 riverine fish communities across France (1995–2018) to evaluate how
biodiversity trends reshape food web structure. We examine relationships among
species richness, community biomass, and three key food web metrics:
connectance, weighted average trophic level, and trophic pathway redundancy. We
also test the space-for-time hypothesis using spatial gradients and a
theoretical food web model. Structural equation models and robustness analyses
reveal that declines in species richness strongly correlate with reduced
community biomass, with top trophic-level species playing a disproportionate
role in food web changes. Declining biomass leads to the loss of top trophic
levels and decreased connectance, while species richness negatively correlates
with connectance. Both biomass and species richness declines reduce trophic
pathway redundancy, suggesting biodiversity loss undermines community robustness
to future perturbations. Our findings align with food web theory and
Biodiversity-Ecosystem Functioning research, support the validity of
space-for-time approaches for basic food web metrics, and highlight the utility
of food web metrics as indicators of ecosystem function and fragility.

# Significance (118 / 120 words max)

Understanding how biodiversity loss affects ecological communities is crucial
for predicting and mitigating human impacts on ecosystems. This study provides
comprehensive evidence on how biodiversity changes are reshaping food web structure using
an extensive dataset of riverine fish communities. We demonstrate that changes
in species richness and community biomass consistently alter trophic levels,
connectance, and the redundancy of feeding relationships. Our findings show how
energetic constraints and biodiversity-ecosystem functioning mechanisms jointly
shape food web assembly and disassembly. Importantly, we show that declining
biodiversity reduces the redundancy of trophic pathways, potentially making
communities more vulnerable to future perturbations. These results suggest that
food web structure metrics could serve as valuable indicators for monitoring
ecosystem health and evaluating conservation strategies.

@@

Human pressures are driving profound biodiversity changes, including freshwater
and terrestrial biodiversity declines linked to land degradation
[@newbold_global_2015; @cardinale_is_2018; @danet_past_2024] and reductions in
marine fish biomass due to intensified fishing pressures on predatory species
[@christensen_century_2014]. A central goal of Biodiversity-Ecosystem
Functioning and food web research has been to understand the ecological
consequences of these widespread biodiversity losses for ecosystem functioning
and stability [@loreau_biodiversity_2001; @loreau_ecological_2022;
@thebault_food-web_2003; @schneider_animal_2016; @dunne_network_2002;
@sole_complexity_2001; @allesina_functional_2009; @montoya_press_2009; @sanders_trophic_2018]. However,
research on biodiversity changes remains largely disconnected from studies
examining their ecological impacts, primarily due to a lack of simultaneous
temporal data on biodiversity and food web dynamics [@danet_species_2021]. As a
result, much of our understanding of the ecological consequences of biodiversity
changes over time relies on experimental and theoretical approaches. These
approaches often depend on the critical assumption that spatial comparisons of
ecological communities can serve as a proxy for temporal trends—a space-for-time
substitution—when direct temporal data are scarce. To advance our understanding
of the ecological consequences of biodiversity changes, it is essential to
investigate how these changes are reshaping food web structure in natural
settings.

Here, we bridge the gap between biodiversity changes and food web dynamics by
leveraging a unique dataset of riverine fish communities sampled repeatedly over
a 23-year period (1995–2018) across 403 locations in France. By reconstructing
yearly food web structures [@danet_species_2021; @bonnaffe_comparison_2021], we
investigate how temporal biodiversity trends reshape food web metrics such as
connectance, weighted average trophic level, and the proportion of redundant
trophic links (hereafter “redundancy”). Linking these findings to established
theories of Biodiversity-Ecosystem Functioning and food web dynamics, we address
three key objectives: (i) documenting how temporal biodiversity trends reshape
food web structure, (ii) connecting these changes to ecosystem functioning and
food web theories, and (iii) assessing the validity of the space-for-time
substitution assumption in the context of biodiversity and food web structure
changes.

The structure of ecological networks, which describes the number, distribution,
and strength of interactions among species, is critical for understanding
species coexistence, ecosystem functioning, and stability to perturbations
[@chesson_interaction_2008; @thebault_food-web_2003; @oconnor_general_2017;
@danet_species_2021; @eschenbrenner_diversity_2023; @danet_response_2024]. Both
theory and empirical evidence demonstrate that species richness and food web
structure influence coexistence [@chesson_interaction_2008], ecosystem
productivity [@thebault_food web_2003; @wang_biodiversity_2018], temporal
stability [@danet_species_2021; @danet_response_2024;
@eschenbrenner_diversity_2023], and robustness to disturbances
[@sole_complexity_2001; @dunne_network_2002; @allesina_functional_2009;
@bodini_using_2009; @montoya_press_2009; @sanders_trophic_2018]. For instance,
keystone predators can regulate dominant competitors, preventing competitive
exclusion and maintaining species richness [@heithaus_predicting_2008;
@johnson_rarity_2007; @estes_trophic_2011]. Similarly, food webs with higher
trophic levels are associated with greater ecosystem productivity
[@wang_biodiversity_2018; @schneider_animal_2016; @wang_biodiversity_2018;
@wang_intraguild_2019; @nie_will_2023] and community stability
[@danet_species_2021; @eschenbrenner_diversity_2023; @danet_response_2024],
while the number and distribution of trophic links determine resilience to
perturbations [@dunne_network_2002; @allesina_who_2004; @bodini_using_2009;
@allesina_functional_2009; @sanders_trophic_2018]. Understanding how
biodiversity changes reshape ecological networks is therefore essential for
quantifying and mitigating anthropogenic impacts on ecosystems.

Three key components link biodiversity changes to food web structure. First, the
energetic paradigm of food web dynamics posits that the presence of top
consumers depends on sufficient resource availability at lower trophic levels,
driven by bottom-up productivity [@lindeman_trophic-dynamic_1942;
@gravel_trophic_2011]. Thus, the biomass of generalist top consumers is likely
constrained by overall community biomass. Second, connectance—the proportion of
realized trophic links—typically decreases with increasing species richness, as
consumers are limited in the number of resources they can exploit
[@oconnor_general_2017; @dunne_network_2006]. However, higher species richness
is expected to increase the redundancy of trophic pathways, enhancing ecosystem
robustness to species extinctions [@dunne_network_2002; @allesina_who_2004;
@bodini_using_2009; @allesina_functional_2009]. This redundancy underpins the
insurance effect of biodiversity, where greater diversity buffers ecosystems
against perturbations and environmental changes [@yachi_biodiversity_1999;
@loreau_ecological_2022]. Finally, biodiversity-ecosystem functioning theory
suggests that species richness promotes community biomass through niche
complementarity [@loreau_biodiversity_2001]. In food webs, the presence of
higher trophic levels reduces niche overlap among producers, further boosting
productivity and biomass [@schneider_animal_2016; @wang_biodiversity_2018;
@wang_intraguild_2019; @nie_will_2023]. Together, these mechanisms highlight how
biodiversity changes can alter trophic link distribution, robustness, and
ecosystem functioning.

To achieve these objectives, we analyzed temporal trends in species richness,
biomass, and food web structure across `r nb_op` community samplings (Fig.
\@ref(fig:matmet)a-c, Table S1). Using structural equation modeling (SEM), we
inferred the relative effects of biomass and species richness trends on food web
metrics, as well as the effect of species richness on biomass (Fig.
\@ref(fig:matmet)d). We also conducted a robustness analysis on the meta-food
web (comprising all trophic species and their interactions) to identify species
extinction scenarios that could explain observed variations in food web
structure. To test the space-for-time hypothesis, we compared temporal findings
with spatial gradients using two approaches: (i) replicating our temporal
analysis using median biodiversity and food web metrics across sites (Fig.
\@ref(fig:matmet)c & e), and (ii) simulating food web dynamics across gradients
of species richness and primary productivity using a bioenergetic model
[@brose_allometric_2006; @lajaaiti_ecologicalnetworksdynamicsjl_2025]. These
simulations allowed us to evaluate whether theoretical predictions align with
our empirical data, which are limited to fish biodiversity.

```{r}
fig_mat_cap <- paste0("Methodological Approach: (a) We analyzed ", nb_st, "
  stream fish communities in France, each monitored for at least 10 years. (b)
  For each site and sampling event, we reconstructed food webs using a metaweb
  based on ontogenetic diet information and predator-prey body size ratios. (c)
  We computed temporal trends and median values for community biomass, species
  richness, connectance, average trophic level, and redundancy at each site. (d)
  We then assessed relationships among these food web descriptors over time and
  (e) across spatial gradients.")
```

@@

Our food web reconstruction delivered `r nb_op` annual estimates of species
richness, biomass, connectance, weighted average trophic level, and the
proportion of redundant trophic links (hereafter “redundancy”, see Appendix B)
from our `r nb_st` fish communities with 10+ years of population dynamics
spanning 1995-2018 over France (Fig. \@ref(fig:matmet)a-c). We implemented Bayesian hierarchical
models to jointly estimate the temporal trends in these metrics. We found a
great heterogeneity in the temporal trends at the site level (community biomass:
from `r grbmn`% to +`r grbmx`% per decade, species richness: `r grrhn` to +`r
grrhx`%, average trophic level: `r grtln` to +`r grtlx`, connectance: `r grctn`
to +`r grctx`, redundancy: `r grln` to +`r grlx`, Fig. S1) which was effectively
captured by the hierarchical model ( conditional R2 ranging from `r
r2cond$log_bm_std` for community biomass to `r r2cond$log_rich_std` for species
richness, Table S2).

```{r}
sem_cap <- "(a) Structural equation models linking the temporal trends of
biodiversity and food web structure. Dashed and solid arrows display
respectively negative and positive the standardized slope coefficients, their
widths being proportional to the absolute values of the coefficients.. The
standardized slope coefficients and their associated 95% confidence intervals
obtained by bootstrapping are displayed along the path. R2 displays the marginal
R-squared of the linear models. (b to h) Direct bivariate relationships derived
from the structural equation model. Points display the temporal trends at each
site estimated with a hierarchical Bayesian model, lines display the linear
relationship estimated with the SEM. The distribution of the site temporal
trends are displayed in Fig. S1. All the effects are reported in Table S3.
Average trophic level is weighted by the biomass of the trophic species.
Redundancy is the proportion of redundant links."

```

@@

## Association between biodiversity and food web temporal trends

These temporal trend estimations at the site level were the input to our SEM,
allowing us to evaluate the direct effect of the temporal trends of species
richness and community biomass on the temporal trends of food web structure, the
direct effect of  species richness temporal trends on the temporal trends of
community biomass, and the indirect effect of species richness temporal trends
on the temporal trends in food web structure through community biomass. We
detected a strong positive effect of species richness temporal trends on the
temporal trends in community biomass ($r_\delta =$ `r rdtps$bm$r`,
Fig. \@ref(fig:sem) a and b, $r_\delta$: standardized slope coefficient, [95%
CI], Table S3). This indicates that communities with decreasing, or increasing,
species richness over time experienced a concomitant decrease, or increase, in
community biomass.

We also find a positive effect of the temporal trend in community biomass on the
temporal trends of connectance, average trophic level, and redundancy (resp.
$r_\delta =$ `r rdtps$ct$bm`, `r rdtps$tlvl$bm`, `r rdtps$red$bm`, Fig.
\@ref(fig:sem)b-d). Thus, when communities experienced a decrease, or increase,
in community biomass, they also experienced a concomitant decrease, or increase,
in connectance, average trophic level and redundancy. To assess whether top
trophic level species were driving those relationships, we additionally looked
at the temporal trends in the number and the proportion of piscivorous species.
The temporal trends in community biomass were positively associated with both
the temporal trends in the number and proportion of piscivorous species (Fig.
S2a). Then, communities with increasing biomass have an increase in average
trophic level, in the number and proportion of piscivorous fishes (or
reciprocally for decreasing biomass), suggesting that temporal trends in
community biomass are driven by high trophic level fishes.

The total effect of the temporal trends in species richness on the temporal
trend in connectance was negative ($r_\delta =$ `r rttps$ct$r`, Fig.
\@ref(fig:std)a), resulting from a negative direct effect ($r_\delta =$
`r rdtps$ct$r`, Fig. \@ref(fig:sem)f) and a positive indirect effect via changes
in the temporal trends in community biomass ($r_\delta =$ `r ridtps$ct$r`).
Thus, when communities experienced an increase, or decrease, in species
richness, they also experienced a concomitant increase, or decrease, in
connectance. In contrast, species richness temporal trends had no direct effects
on trends in average trophic level ($r_\delta =$ `r rdtps$tlvl$r`, Fig.
\@ref(fig:sem)g), but a positive indirect effect via changes in the trends of
community biomass ($r_\delta =$ `r ridtps$tlvl$r`), which
adds up to a total positive effect of species richness temporal trends on
average trophic level ($r_\delta =$ `r rttps$tlvl$r`, Fig. \@ref(fig:std)a).
Temporal trends in species richness had a strong total positive effect on the
trends in redundancy (resp. $r_\delta =$ `r rdtps$red$r`, `r ridtps$red$r`), via
direct and indirect positive effects ($r_\delta =$ `r rttps$red$r`). All
together, it means that communities that have experienced species richness
declines over the last decades have decreased in average trophic level and
redundancy but increased in connectance. Conversely, those that experienced
species richness increases have increased in average trophic level, redundancy,
but decreased in connectance.

The effects of species richness were more complex, because species richness in
this model has indirect effects on food web structure through community biomass.
Species richness had a direct negative effect on connectance ($r_\delta =$
`r rdtps$ct$r`, Fig. \@ref(fig:sem)f), a positive indirect effect through
community biomass ($r_\delta =$ `r ridtps$ct$r`), so that species richness had a
negative effect in total on connectance ($r_\delta =$ `r rttps$ct$r`, Fig.
\@ref(fig:std)a). In contrast, species richness had no direct effects on in
average trophic level ($r_\delta =$ `r rdtps$tlvl$r`, Fig. \@ref(fig:sem)g), a
positive indirect effect through community biomass ($r_\delta =$
`r ridtps$tlvl$r`), which adds up to a positive total effect of species richness
temporal trends on average trophic level ($r_\delta =$ `r rttps$tlvl$r`, Fig.
\@ref(fig:std)a). Species richness temporal trends had direct and indirect
positive effect on the temporal trends in redundancy (resp. $r_\delta =$
`r rdtps$red$r`, `r ridtps$red$r`), which added up to a strong positive effect
($r_\delta =$ `r rttps$red$r`).

Spatial analyses, either using empirical data or the bioenergetic food web
model, largely aligned with temporal trends (Fig. \@ref(fig:std)a and S3),
showing consistent positive effects of species richness on community biomass and
of community biomass on average trophic level and redundancy. Species richness
also exhibited a direct negative effect on connectance and a positive effect on
redundancy, mirroring temporal patterns. However, spatial and temporal analyses
diverged in their effects of species richness on average trophic level, with
spatial data showing a negative effect and temporal trends indicating a positive
one. Detailed statistical effects and comparisons are provided in the Appendix.

@@

We implemented three secondary extinction simulation scenarios on the empirical
metaweb defined as a food web containing all the trophic species, i.e. the fish
species subdivided in size classes added to their resources, found across all
sites in the empirical data. This was to access a second line of inference about
whether top species were drivers of biomass and structure patterns (see section
4.2). When species are removed in a sequential order beginning by the highest
trophic levels, connectance, average trophic level, and redundancy decreased
(purple lines, Fig. \@ref(fig:std)b). On the other hand, removing low trophic level species
first increases connectance and average trophic level, but did not affect
redundancy (yellow lines, Fig. \@ref(fig:std)b). In turn, random species extinction sequence
had little effect on food web structure (black line, Fig. \@ref(fig:std)b). These results,
in conjunction with our empirical analyses above, suggest that top consumers
play a disproportionate role in driving the changes in food web structure
reported.

```{r}
std_cap <- paste0( "a, Direct and total effects of community biomass and species
  richness trends on food web structure trends. Standardized coefficients (r)
  derived from the temporal, spatial, and bioenergetic model Structural Equation
  Models. Points and lines display the estimated effects and their confidence
  intervals at 95\\%. Values are displayed in Table S3, R-squared in Table S4, and
  the visualization of the bivariate relationships for the spatial and
  bioenergetic model SEMs are displayed in Figure S3. Average trophic level is
  weighted by the biomass of the trophic species. Redundancy is the proportion
  of redundant links. b, Effect of species extinctions on the structure of the
  empirical metaweb according to three scenarios. In the Decreasing and
  Increasing scenarios, species were ordered by their trophic level and removed
  from the metaweb starting respectively from the highest and lowest trophic
  level. Species were removed in a random sequence (50 sequences, grey dots) for
  the random scenario, the black line displaying the average predicted by a GAM
  linear model. For the extinction sequence, the average trophic level was not
  weighted by the biomass of species as this information is not available at the
  metaweb scale. Similar results were obtained with trophic species (i.e. node)
  extinction (Fig. S4).")
```
